---
title: "Chapter 1 - EDA"
subtitle: "Italian Oils: a study of 572 different oils based on eight fatty acids"
author: "Davide Carrieri"
date: today
---


## The question

Italy's biodiversity is incredibly rich for a country of its size, a fact that is highly evident in its olive oil. Regional differences across the country drastically alter the oil's taste, color, and quality.

The aim of this study is to analyze the fatty acid composition of 572 olive oil samples to characterize their geographic origin and understand how different areas affect the final product.

## Getting the data

The data "olive", the one that is going to be studied, come from an R package called "dslabs". 

The dataset counts 572 observations and 10 variables:

  - region. General region of Italy (Northern Italy, Southern Italy and Sardinia)

  - area. Area of Italy (Calabria, Coast Sardinia,East Liguria, Inland Sardinia, North Apulia, Sicily, South Apulia, Umbria, West Liguria)

  - palmitic. Percent palmitic acid of sample.

  - palmitoleic. Percent palmitoleic of sample.

  - stearic. Percent stearic acid of sample.

  - oleic. Percent oleic acid of sample.

  - linoleic. Percent linoleic acid of sample.

  - linolenic. Percent linolenic acid of sample.

  - arachidic. Percent arachidic acid of sample.

  - eicosenoic. Percent eicosenoic acid of sample.

```{r}
#| label: setup
#| message: false

library(dslabs)
library(janitor)        
library(naniar)         
library(skimr)         
```




## A first look
```{r}
#| label: download
#| message: false

data <- dslabs::olive
summary(data)
str(data)
```
```{r}
#| label: skim
#| fig-cap: "A summary of the dataset."
# The skim() output gives you type, missingness and distributions at a glance.
skim(data)
```

```{r}
#| label: missingness
#| fig-cap: "Missing values across variables."
 vis_miss(data) #— where is data missing, and why might that be?
```

## Cleaning decisions

Since there are 0 missing values, the choice is to the preserve all data, as they are all part of our object of studies.

## Plotting the story

<!-- TODO: replace with your figures. Give every figure a caption and refer to
it in the text (e.g. "Fig. 1 shows that …"). If a figure uses randomness
(sampling, jitter, a simulation), fix the seed *inside* that cell with
`withr::with_seed(2025, { ... })` so the figure is identical on every render. -->

### 1. Boxplot
```{r}
#| label: fig-all-acids-boxplot
#| fig-cap: "Distribution of all fatty acids by geographic area."
#| fig-width: 10
#| fig-height: 18
#| warning: false
#| echo: false

library(tidyverse)

data %>%
  pivot_longer(cols = palmitic:eicosenoic, 
               names_to = "acid", 
               values_to = "percentage") %>%
  ggplot(aes(x = area, y = percentage, fill = area)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~ acid, scales = "free_y", ncol = 2) +
  scale_x_discrete(sec.axis = dup_axis(name = NULL)) +
  theme_minimal() +
  theme(
    
    axis.text.x.bottom = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.text.x.top = element_text(angle = 90, hjust = 0, vjust = 0.5),
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 12)
  ) +
  labs(x = NULL, y = "Concentration (%)")
```

### 2. Heatmap

```{r}
#| label: fig-heatmap-acids
#| fig-cap: "Chemical profile by area (normalized values compared to the mean of each acid)."
#| echo: false

data %>%
  # 1. Calculate the mean of all acids grouped by area
  group_by(area) %>%
  summarise(across(palmitic:eicosenoic, mean)) %>%
  # 2. Convert to long format
  pivot_longer(cols = -area, names_to = "acid", values_to = "mean_val") %>%
  # 3. Normalize the values for each individual acid (Z-score)
  group_by(acid) %>%
  mutate(scaled_val = scale(mean_val)) %>%
  # 4. Create the heatmap
  ggplot(aes(x = acid, y = area, fill = scaled_val)) +
  geom_tile(color = "white", linewidth = 0.5) +
  # Use a diverging color palette (blue = below average, red = above average)
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick", 
                       midpoint = 0, name = "Deviation\nfrom mean") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Fatty Acid", y = "Area")
```
## Findings

<!-- TODO: what did you learn? This is your discussion: put your findings and
the original question back into perspective. Keep it honest: say what you
found, and what you did NOT find. -->

---

*This chapter is your semester's centrepiece. Revisit it as you learn new
techniques and make it your best work by the final gallery walk.*
